WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-09-23 ms.tex
Adjustments for post-O1 submission

The \acl{LIGO} (\acsu{LIGO}; \cite{AdvancedLIGOHarry,AdvancedLIGO}) and Virgo \cite{AdvancedVirgo} will soon begin taking data \cite{LIGOObservingScenarios} in their `Advanced' configurations, ultimately increasing their reach in volume within the local Universe by three orders of magnitude as compared to their initial configurations through 2010. They will form the first parts of a sensitive global \ac{GW} detector network, to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India \cite{LIGOIndia}.

@@

We call this algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} will provide localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a key role in enabling rapid follow-up observations.

@@

One immediately pressing direction for future work is to address the issue of sub\nobreakdashes-threshold signals, as this will be a major issue when Advanced Virgo comes online in 2016\nobreakdashes--2017. Using the full \ac{SNR} time series in place of the autocorrelation function seems like a promising avenue; implementing this requires some infrastructure changes to both the matched filter pipeline and \ac{BAYESTAR}. Along these lines, we refer the reader to \cite{PankowRapidPE} for a similar, non-\ac{MCMC} approach to rapid exploration of the full parameter space.

A more open question is how to account for spin precession. The simulations in \cite{FirstTwoYears} and in this paper featured extremely modest spins of $\chi \leq 0.05$, consistent with the fastest known pulsars in binaries \citep{2003Natur.426..531B,DetectingBNSSystemsWithSpin}. The signals were detected using a template bank that lacked spins entirely. Farr et~al. (in preparation) show that using nonspinning \ac{BNS} templates for parameter estimation has negligible impact on sky localization. However, if one or both companions are spinning as fast as a millisecond pulsar, $\chi \sim 0.4$ \cite{FastestSpinningMillisecondPulsar}, or even near breakup, $\chi \sim 0.7$, then the orbital plane may precess; in this case, spins can no longer be neglected for detection \cite{DetectingBNSSystemsWithSpin} and may also be important for parameter estimation. Since spinning \ac{BNS} searches are still an active area of development, \ac{BAYESTAR}'s sky localization accuracy in this regime should be reexamined in the future.

@@

On the topic of very low\nobreakdashes-latency localization, we also recommend Chen~\&~Holz (in preparation), who propose a rapid localization scheme that is similar to ours, but even faster: their likelihood is strictly Gaussian, so one more marginalization integral (the integral over arrival time) can be performed analytically.

>>>>

The \acl{LIGO} (\acsu{LIGO}; \cite{AdvancedLIGOHarry,AdvancedLIGO}) and Virgo \cite{AdvancedVirgo} have just begun taking data \cite{LIGOObservingScenarios} in their `Advanced' configurations. They will ultimately increase their reach in volume within the local Universe by three orders of magnitude as compared to their initial configurations through 2010. They form the first parts of a sensitive global \ac{GW} detector network, to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India \cite{LIGOIndia}.

@@

We call this algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} is providing localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a key role in enabling rapid follow-up observations.

@@

One immediately pressing direction for future work is to address the issue of sub\nobreakdashes-threshold signals, as this will be a major issue when Advanced Virgo comes online in 2016\nobreakdashes--2017. Using the full \ac{SNR} time series in place of the autocorrelation function seems like a promising avenue; implementing this requires some infrastructure changes to both the matched filter pipeline and \ac{BAYESTAR}. Along these lines, we also refer the reader to \cite{PankowRapidPE} for a similar, non-\ac{MCMC} approach to rapid exploration of the full parameter space.

A more open\nobreakdashes-ended question is how to account for spin precession. The simulations in \cite{FirstTwoYears} and in this paper featured extremely modest spins of $\chi \leq 0.05$, consistent with the fastest known pulsars in binaries \citep{2003Natur.426..531B,DetectingBNSSystemsWithSpin}. The signals were detected using a template bank that lacked spins entirely. Farr et~al. (in preparation) show that using nonspinning \ac{BNS} templates for parameter estimation has negligible impact on sky localization. However, if one or both companions are spinning as fast as a millisecond pulsar, $\chi \sim 0.4$ \cite{FastestSpinningMillisecondPulsar}, or even near breakup, $\chi \sim 0.7$, then the orbital plane may precess; in this case, spins can no longer be neglected for detection \cite{DetectingBNSSystemsWithSpin} and may also be important for parameter estimation. Since spinning \ac{BNS} searches are still an active area of development, \ac{BAYESTAR}'s sky localization accuracy in this regime should be reexamined in the future.

@@

On the topic of very low\nobreakdashes-latency localization, we also recommend Chen~\&~Holz (in preparation), who propose a rapid localization scheme that is similar to ours, but even faster because it makes some additional compromises: their likelihood is strictly Gaussian, so one more marginalization integral (the integral over arrival time) can be performed analytically.
